A Framework for Modelling Thin Objects: \\ from Meshing to Shell Finite Element Method

### 2010-03-09 MICCAI2010/paper.tex
comments
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@86 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\usepackage[english]{babel}
%\usepackage[tight]{subfigure}
%\usepackage{floatrow}
\usepackage{graphicx}
%\usepackage{amsmath}
%\usepackage{multirow}
%\renewcommand*\arraystretch{1.25} % extra vertical space in cells
\usepackage{amsmath}

@@

Human body is composed of various deformable anatomical structures. 
Realistic modeling of organ deformations is a challenging research field that leads the way to numerous clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation.
In all these cases, the clinician needs fast update of the deformation model to obtain a real-time display of the computed deformations.  
To realize this kind of interactive systems, one face the trade-off problem between accuracy of the model and the need of fast computations.

An other key challenge of real-time human soft-tissue modeling is the variousness of the mechanical behaviors. 
It seems unrealistic to target a unique model for all tissues even if there are obvious advantages to make maximum use of optimized generic models.
Yet most of previous work focus on volume models that are able to capture the behavior of volume organs like liver \cite{} or brain tissues \cite{}.
In contrast, this paper seeks to propose a solution for simulating, in real-time, deformation of thin anatomical structures, whose volume is negligible compared to their surface area.
Examples of such anatomical structures include hollow structures, like the wall of blood vessels or the colon but also the envelope of volume organs, like the capsule of Glisson.

@@

Development of a satisfactory physical model that runs in real-time but produces visually convincing animation of thin objects is challenging. Rather than resorting to shell theory which involves the most complex formulations in continuum mechanics, previous works have often relied on discrete formulations. Early approaches only considered in-plane deformation, and often relied on mass-spring models. More recent works have considered adding bending through angular springs. For instance Wang et al.~\cite{Wang07} successfully used a network of linear and angular springs to describe bending and twisting of catheters and guidewires in an interventional radiology simulator. Yet, such models are limited in their ability to describe certain behaviour, as they do not rely on continuum mechanics. Another limitation of such models is the difficulty to derive spring stiffness (in particular for angular springs) from elastic properties (Young's modulus and Poisson's ratio). For these reasons, other approaches have been proposed.

Continuum mechanics provides many formulations able to accurately describe stresses occurring within thin objects. Most of them fall into one of the following two categories: plate theory or shell theory. Those theories have been a subject of interest in the mechanical community for decades. The difference between these two kinds of structures is very well explained by Liu et al.~\cite{Liu03} and can be summarized by the fact that plate bending elements can only carry transversal loads while shells can undergo more complex deformations. Considering the wide range of shapes and solicitations that can be applied to hollow organs, a thin plate model would not correctly capture their deformation and therefore a shell formulation was retained.

@@

A complete description of our co-rotational triangular shell finite element model is available in one of our previous publication \cite{Comas2010ISBMS}. Therefore we will only remind briefly the key points. We improved and extended a plate model first introduced by Przemieniecki \cite{Przemieniecki68} to a co-rotational formulation. Co-rotational approaches offer a good trade-off between computational efficiency and accuracy by allowing small deformations but large displacements. Once combined with an in-plane membrane formulation we obtain an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies.

@@

We can calculate the strains from the flat-plate theory using:
\begin{equation}
e_{xx} = -z \frac{\partial^2u_z}{\partial x^2}
\end{equation} 
\begin{equation}
e_{yy} = -z \frac{\partial^2u_z}{\partial y^2}
\end{equation} 
\begin{equation}
e_{xy} = -2z \frac{\partial^2u_z}{\partial x \partial y}
\end{equation}

Symbolically this may be expressed as $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ derives from equation~\ref{eq-deflection} and the above relations. Noting that $\textbf{c} = \textbf{C}^{-1}\textbf{u}$, we have:
\begin{equation}
\textbf{e} = \textbf{DC}^{-1}\textbf{u} = \textbf{bu}
\end{equation} 
where the strain-displacement matrix $\textbf{b} = \textbf{DC}^{-1}$. 
The stiffness matrix $\textbf{K}_e$ for an element is then classically obtained from
\begin{equation}
\textbf{K}_e = \int_v \textbf{b}^{T} \boldsymbol\chi \textbf{b} dV
\end{equation} 
where $\boldsymbol\chi$ is the material matrix. The stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element as in the membrane case.

@@

\begin{equation}
d_{\mathrm{H}}(X,Y) = \max \left\{ \sup_{x \in X} \inf_{y \in Y} d(x,y), \sup_{y \in Y} \inf_{x \in X} d(x,y) \right\} \enspace .
\end{equation}
where $d()$ is the Euclidian distance between two points. Since the curved surface of each shell must be taken into account in the computation of the Hausdorff distance, a finer mesh is created by adding vertices on the surface of each element. We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using (\ref{eq-deflection}). We then use this new mesh that samples the actual surface described by the shells to compute the Hausdorff distance with the targeted high resolution mesh. The same process of subdivision allows us to render each shell as a curved triangle (Fig.~\ref{fig-shell} ).

@@

The first step in the process of generating a shell-based mesh is an aggressive decimation of the targeted high resolution mesh. We then apply a heuristic method derived from the work of Saupin et. al \cite{Saupin07} with tetrahedral meshes based on very simple geometrical rules. For each node of the coarse mesh, we find the three closest triangles on the high resolution mesh and we move the node to the barycenter of the three centres of mass of those triangles. This technique locally smooths the surface of the mesh while converging towards the desired high resolution mesh. At each iteration of this algorithm we measure the error with the target using the Hausdorff distance and the process is stopped when the required precision has been reached. A simple example is shown Fig. \ref{fig-cylinder} and \ref{fig-cylinderResults} to illustrate the method.

>>>>

\usepackage[english]{babel}
%\usepackage[tight]{subfigure}
%\usepackage{floatrow}
\usepackage{graphicx}
%\usepackage{amsmath}
%\usepackage{multirow}
%\renewcommand*\arraystretch{1.25} % extra vertical space in cells
\usepackage{amsmath}
\usepackage[latin1]{inputenc} % pour les caracteres francais dans les remarques

\usepackage[usenames]{color}
\newcommand{\SC}[1]{{\color{red}{\textbf{SC: #1}}}}
\newcommand{\OC}[1]{{\color{blue}{\textbf{OC: #1}}}}
\newcommand{\CD}[1]{{\color{magenta}{\textbf{CD: #1}}}}

\newcommand{\nomore}[1]{}

@@

Human body is composed of various deformable anatomical structures. 
Realistic modeling of organ deformations is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation.
In all these cases, the clinician needs fast update of the deformation model to obtain a real-time display of the computed deformations.  
To realize this kind of interactive systems, one face the trade-off problem between accuracy of the model and the need of fast computations.

An other key challenge of real-time human soft-tissue modeling is the variousness of the mechanical behaviors. 
It seems unrealistic to target a unique model for all tissues even if there are obvious advantages to make maximum use of optimized generic models.
Yet most of previous work focus on volume models that are able to capture the behavior of volume organs like liver \cite{} or brain tissues \cite{} \CD{TODO: biblio}.
In contrast, this paper seeks to propose a solution for simulating, in real-time, deformation of thin anatomical structures, whose volume is negligible compared to their surface area.
Examples of such anatomical structures include hollow structures, like the wall of blood vessels or the colon but also the envelope of volume organs, like the capsule of Glisson.

@@

Development of a satisfactory physical model that runs in real-time but produces visually convincing animation of thin objects is challenging. \CD{utile?}
Rather than resorting to shell theory which involves the most complex formulations in continuum mechanics, previous works have often relied on discrete formulations. Early approaches only considered in-plane deformation, and often relied on mass-spring models. More recent works have considered adding bending through angular springs. For instance Wang et al.~\cite{Wang07} successfully used a network of linear and angular springs to describe bending and twisting of catheters and guidewires in an interventional radiology simulator. \CD{C'est pas des shells si?} Yet, such models are limited in their ability to describe certain behaviour, as they do not rely on continuum mechanics. Another limitation of such models is the difficulty to derive spring stiffness (in particular for angular springs) from elastic properties (Young's modulus and Poisson's ratio). For these reasons, other approaches have been proposed.

Continuum mechanics provides many formulations able to accurately describe stresses occurring within thin objects. Most of them fall into one of the following two categories: plate theory or shell theory. Those theories have been a subject of interest in the mechanical community for decades. The difference between these two kinds of structures is very well explained by Liu et al.~\cite{Liu03} and can be summarized by the fact that plate bending elements can only carry transversal loads while shells can undergo more complex deformations. Considering the wide range of shapes and solicitations that can be applied to hollow organs, a thin plate model would not correctly capture their deformation and therefore a shell formulation was retained for this work.

@@

A complete description of our co-rotational triangular shell finite element model is available in one of our previous publication \cite{Comas2010ISBMS} \CD{Attention, il faut faire une citation anonyme !!}. Therefore we will only remind briefly the key points. We improved and extended a plate model first introduced by Przemieniecki \cite{Przemieniecki68} to a co-rotational formulation. Co-rotational approaches offer a good trade-off between computational efficiency and accuracy by allowing small deformations but large displacements. Once combined with an in-plane membrane formulation we obtain an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies.

@@

We can calculate the strains from the flat-plate theory using:
\begin{equation}
\label{eq-deformation}
e_{xx} = -z \frac{\partial^2u_z}{\partial x^2}
\hspace{1cm}
e_{yy} = -z \frac{\partial^2u_z}{\partial y^2}
\hspace{1cm}
e_{xy} = -2z \frac{\partial^2u_z}{\partial x \partial y}
\end{equation}

Symbolically this may be expressed as $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ derives from equations~\ref{eq-deflection} and~\ref{eq-deformation}. Noting that $\textbf{c} = \textbf{C}^{-1}\textbf{u}$, we have:
\begin{equation}
\textbf{e} = \textbf{DC}^{-1}\textbf{u} = \textbf{bu}
\end{equation} 
where the strain-displacement matrix $\textbf{b} = \textbf{DC}^{-1}$. 
The stiffness matrix $\textbf{K}_e$ for an element is then classically obtained from
\begin{equation}
\textbf{K}_e = \int_v \textbf{b}^{T} \boldsymbol\chi \textbf{b} dV
\end{equation} 
where $\boldsymbol\chi$ is the material matrix. The stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element as in the membrane case.

@@

\begin{equation}
d_{\mathrm{H}}(X,Y) = \max \left\{ \sup_{x \in X} \inf_{y \in Y} d(x,y), \sup_{y \in Y} \inf_{x \in X} d(x,y) \right\} \enspace .
\end{equation}
where $d()$ is the Euclidian distance between two points. Since the curved surface of each shell must be taken into account in the computation of the Hausdorff distance, a finer mesh is created by adding vertices on the surface of each element. We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using (\ref{eq-deflection}). \CD{Lien avec la fonction d'interpolation a mettre plus en avant !}. We then use this new mesh that samples the actual surface described by the shells to compute the Hausdorff distance with the targeted high resolution mesh. The same process of subdivision allows us to render each shell as a curved triangle (Fig.~\ref{fig-shell} ).

@@

The first step in the process of generating a shell-based mesh is an aggressive decimation of the targeted high resolution mesh. We then apply a heuristi"c method derived from the work of Saupin et. al \cite{Saupin07} with tetrahedral meshes based on very simple geometrical rules. For each node of the coarse mesh, we find the three closest triangles on the high resolution mesh and we move the node to the barycenter of the three centres of mass of those triangles. This technique locally smooths the surface of the mesh while converging towards the desired high resolution mesh. At each iteration of this algorithm we measure the error with the target using the Hausdorff distance and the process is stopped when the required precision has been reached. A simple example is shown Fig. \ref{fig-cylinder} and \ref{fig-cylinderResults} to illustrate the method.
